---
title: "Income and Cash Flow Statments"
output: beamer_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Goals

+ Make what you learned about taxes and depreciation understandable to others.
+ Use accounting conventions to convey your cash flow projections.

## Background

+ We will be focusing on the cash flow effects of some investment.
+ We will not be looking at the whole business, just the thing that changed.
+ We are looking at a counter-factual, "How would cash flow change?"

## The Example Counter-factual

+ How would cash flow change if we purchased a new inventory management system that reduces our need for inventory?
+ How would cash flow change if we purchased a new delivery truck with a loan?  
+ Is cash flow better than without the loan?

## Sign Conventions Revenue


\begin{table}[h]
\begin{tabular}{l||c|c||p{3cm}}
\multicolumn{4}{c}{Income Statement}\\
  &Year 0  &Year 1  &Note\\
\hline
		Operating Revenue	&	100 & (500)	&\\
		Operating Expenses	&	&	&\\
		Interest	&	&	&\\
		Depreciation	&	&	&\\
        \hline
		Taxable Income	&	&	&\\
		Tax	&	&	&\\
        \hline
		Net Income	&	&	&\\
\end{tabular}
\end{table}

More revenue than current plan in year zero and less revenue than current plan in year one.

## Sign Conventions Expenses


\begin{table}[h]
\begin{tabular}{l||c|c||p{3cm}}
\multicolumn{4}{c}{Income Statement}\\
  &Year 0  &Year 1  &Note\\
\hline
		Operating Revenue	&	100 & (500)	&\\
		Operating Expenses	&	50& (700)	&\\
		Interest	&	&	&\\
		Depreciation	&	&	&\\
        \hline
		Taxable Income	&	&	&\\
		Tax	&	&	&\\
        \hline
		Net Income	&	&	&\\
\end{tabular}
\end{table}

More expenses than current plan in year zero and fewer expenses than current plan in year one.



## Sign Conventions Interest


\begin{table}[h]
\begin{tabular}{l||c|c||p{3cm}}
\multicolumn{4}{c}{Income Statement}\\
  &Year 0  &Year 1  &Note\\
\hline
		Operating Revenue	&	100 & (500)	&\\
		Operating Expenses	&	50& (700)	&\\
		Interest	&	&	300&\\
		Depreciation	&	&	&\\
        \hline
		Taxable Income	&	&	&\\
		Tax	&	&	&\\
        \hline
		Net Income	&	&	&\\
\end{tabular}
\end{table}

Increase in interest expense, relative to current plan, in year 1.  Note that this is just the interest expense, not a full loan payment.


## Sign Conventions Depreciation


\begin{table}[h]
\begin{tabular}{l||c|c||p{3cm}}
\multicolumn{4}{c}{Income Statement}\\
  &Year 0  &Year 1  &Note\\
\hline
		Operating Revenue	&	100 & (500)	&\\
		Operating Expenses	&	50& (700)	&\\
		Interest	&	&	300&\\
		Depreciation	&	4000&	3200&\\
        \hline
		Taxable Income	&	&	&\\
		Tax	&	&	&\\
        \hline
		Net Income	&	&	&\\
\end{tabular}
\end{table}

If you replaced an asset with one that cost less, and had less depreciation in years zero and one, then you would see negative values.


## Taxable Income

Subtract your expenses from your revenue.

\begin{table}[h]
\begin{tabular}{l||c|c||p{3cm}}
\multicolumn{4}{c}{Income Statement}\\
  &Year 0  &Year 1  &Note\\
\hline
		Operating Revenue	&	100 & (500)	&\\
		Operating Expenses	&	50& (700)	&\\
		Interest	&	&	300&\\
		Depreciation	&	4000&	3200&\\
        \hline
		Taxable Income	&	(`r -(100 - 50 -4000)`)& (`r -(-500 +700 -3000 -3200)`)	&\\
		Tax	&	&	&\\
        \hline
		Net Income	&	&	&\\
\end{tabular}
\end{table}


Year 0 Taxable Income = 100 -50 -4000

Year 1 Taxable Income = -5000 - (-700) -300 - 3200

## Tax

Apply your tax rate (40\%) to the Taxable Income. Note that increases in taxes are positive but decreases in taxes are shown as a negative.

\begin{table}[h]
\begin{tabular}{l||c|c||p{3cm}}
\multicolumn{4}{c}{Income Statement}\\
  &Year 0  &Year 1  &Note\\
\hline
		Operating Revenue	&	100 & (500)	&\\
		Operating Expenses	&	50& (700)	&\\
		Interest	&	&	300&\\
		Depreciation	&	4000&	3200&\\
        \hline
		Taxable Income	&	(`r -(100 - 50 -4000)`)& (`r -(-500 +700 -3000 -3200)`)	&\\
				Tax	&	(`r -(100 - 50 -4000)*.4`)&	(`r -(-500 +700 -3000 -3200)*.4`)&\\
        \hline
		Net Income	&	&	&\\
\end{tabular}
\end{table}

This says that your taxes will be *lower* because of the action you are taking.

## Net Income

Subtract Tax from Taxable Income to get Net Income, i.e., profits.

\begin{table}[h]
\begin{tabular}{l||c|c||p{3cm}}
\multicolumn{4}{c}{Income Statement}\\
  &Year 0  &Year 1  &Note\\
\hline
		Operating Revenue	&	100 & (500)	&\\
		Operating Expenses	&	50& (700)	&\\
		Interest	&	&	300&\\
		Depreciation	&	4000&	3200&\\
        \hline
		Taxable Income	&	(`r -(100 - 50 -4000)`)& (`r -(-500 +700 -3000 -3200)`)	&\\
				Tax	&	(`r -(100 - 50 -4000)*.4`)&	(`r -(-500 +700 -3000 -3200)*.4`)&\\
        \hline
		Net Income &	(`r -(100 - 50 -4000)*.6`)&	(`r -(-500 +700 -3000 -3200)*.6`)&\\
\end{tabular}
\end{table}

This says that our profits are lower than they would be otherwise.

## Cash Flow

Cash from operations is just copied from the income statement

\begin{table}
\begin{tabular}{l||c|c||p{3cm}}
        \multicolumn{4}{c}{Cash Flow Statement}\\
  &Year 0  &Year 1  &Note\\        
	Operations&	&	&\\
	\hspace{.25in}Net Income	&	(`r NI0 <- -(100 - 50 -4000)*.6; NI0`)&	(`r NI1 <- -(-500 +700 -3000 -6400)*.6; NI1`)&\\
	\hspace{.25in}Depreciation	&	4000&	3200&\\
	Investments	&	&	&\\
			&	&	&\\
\hspace{.25in}Working Capital	&	&	&\\
		\hspace{.25in}Gains Tax	&	&	&\\
			&	&	&\\
        
	Finance	&	&	&\\
			&	&	&\\
        \hline
		Net Cash Flow	&	&	&\\
\end{tabular}
\end{table}

Keep in mind that the convention is cash inflows are positive and cash outflows are negative.



## Asset Purchase

The Asset is a 20K, 5-Year Asset

\begin{table}
\begin{tabular}{l||c|c||p{3cm}}
        \multicolumn{4}{c}{Cash Flow Statement}\\
  &Year 0  &Year 1  &Note\\        
	Operations&	&	&\\
	\hspace{.25in}Net Income	&	(`r NI0`)&	(`r NI1`)&\\
	\hspace{.25in}Depreciation	&	4000&	3200&\\
	Investments	&	&	&\\
\hspace{.25in}The Asset			&(20000)	&	&\\
\hspace{.25in}Working Capital	&	&	&\\
		\hspace{.25in}Gains Tax	&	&	&\\
			&	&	&\\
        
	Finance	&	&	&\\
			&	&	&\\
        \hline
		Net Cash Flow	&	&	&\\
\end{tabular}
\end{table}

Outflows of cash are negative. Note depreciation reflects sale in year 1.


## Asset Sale

Sell the asset in year 1 for 19000

\begin{table}
\begin{tabular}{l||c|c||p{3cm}}
        \multicolumn{4}{c}{Cash Flow Statement}\\
  &Year 0  &Year 1  &Note\\        
	Operations&	&	&\\
	\hspace{.25in}Net Income	&	(`r NI0`)&	(`r NI1`)&\\
	\hspace{.25in}Depreciation	&	4000&	3200&\\
	Investments	&	&	&\\
\hspace{.25in}The Asset			&(20000)	&	19000&\\
\hspace{.25in}Working Capital	&	&	&\\
		\hspace{.25in}Gains Tax	&	&	&\\
			&	&	&\\
        
	Finance	&	&	&\\
			&	&	&\\
        \hline
		Net Cash Flow	&	&	&\\
\end{tabular}
\end{table}

Inflows of cash are positive.


## Asset Sale

```{r, message=FALSE, warning=FALSE, include=FALSE}
Book <- 20000 - 4000 - 3200
Gains <- (19000 - Book) * .4
```


Gains tax has opposite sign convention.  Positive means pay.  

\begin{table}
\begin{tabular}{l||c|c||p{3cm}}
        \multicolumn{4}{c}{Cash Flow Statement}\\
  &Year 0  &Year 1  &Note\\        
	Operations&	&	&\\
	\hspace{.25in}Net Income	&	(`r NI0`)&	(`r NI1`)&\\
	\hspace{.25in}Depreciation	&	4000&	3200&\\
	Investments	&	&	&\\
\hspace{.25in}The Asset			&(20000)	&	19000&\\
\hspace{.25in}Working Capital	&	&	&\\
		\hspace{.25in}Gains Tax	&	&	&\\
			&	&	&\\
        
	Finance	&	&	&\\
			&	&	&\\
        \hline
		Net Cash Flow	&	&	&\\
\end{tabular}
\end{table}

Book Value = 20000 - 4000 - 3200 = 12800


## Gains Tax

```{r, message=FALSE, warning=FALSE, include=FALSE}
Book <- 20000 - 4000 - 3200
Gains <- (19000 - Book) * .4
```


Gains tax has opposite sign convention.  Positive means pay.  

\begin{table}
\begin{tabular}{l||c|c||p{3cm}}
        \multicolumn{4}{c}{Cash Flow Statement}\\
  &Year 0  &Year 1  &Note\\        
	Operations&	&	&\\
	\hspace{.25in}Net Income	&	(`r NI0`)&	(`r NI1`)&\\
	\hspace{.25in}Depreciation	&	4000&	3200&\\
	Investments	&	&	&\\
\hspace{.25in}The Asset			&(20000)	&	19000&\\
\hspace{.25in}Working Capital	&	&	&\\
		\hspace{.25in}Gains Tax	&	&`r Gains`	&\\
			&	&	&\\
        
	Finance	&	&	&\\
			&	&	&\\
        \hline
		Net Cash Flow	&	&	&\\
\end{tabular}
\end{table}

$Gains~Tax = (Salvage - Book~Value) Tax~Rate =$


## Net Cash Flow

```{r, message=FALSE, warning=FALSE, include=FALSE}

Cash1 <- -2370 + 4000 - 20000
Cash2 <- -5520 + 3200 + 19000 - 2480

```

Add up cash but remember gains tax must be subtracted.

\begin{table}
\begin{tabular}{l||c|c||p{3cm}}
        \multicolumn{4}{c}{Cash Flow Statement}\\
  &Year 0  &Year 1  &Note\\        
	Operations&	&	&\\
	\hspace{.25in}Net Income	&	(`r NI0`)&	(`r NI1`)&\\
	\hspace{.25in}Depreciation	&	4000&	3200&\\
	Investments	&	&	&\\
\hspace{.25in}The Asset			&(20000)	&	19000&\\
\hspace{.25in}Working Capital	&	&	&\\
		\hspace{.25in}Gains Tax	&	&`r Gains`	&\\
			&	&	&\\
        
	Finance	&	&	&\\
			&	&	&\\
        \hline
		Net Cash Flow	&	(18370)& 14200	&\\
\end{tabular}
\end{table}


## Two More Conventions

We still need to show how to treat loans and working capital.

+ Loans
    + Funding recorded on cash flow
    + Payments are split into interest expense (Income Statement) and principal payment (Cash Flow Statement)
+ Working Capital Changes
    + Changes in inventory level  and changes in average accounts receivable.
    + Record how the level changes from period to period.
    

## From the Loans Slides

```{r include=FALSE}
FinFormat <- function(x) format(x, digits=2, decimal.mark=".",   big.mark=",", small.interval=3)


Acc<-function(x){format(x,nsmall=2,digits=2,big.mark=',',scientific=FALSE)}
pgivena<-function(a,i,n){a*((1+i)^n -1)/(i*(1+i)^n)}
pgiveng<-function(g,i,n){g*( (1+i)^n -(i*n) -1  )/(i^2*(1+i)^n)}
agivenp<-function(p,i,n){p*(i*(1+i)^n)/((1+i)^n -1)}
pgivengrow<-function(a1,g,i,n){a1*(1-(1+g)^n*(1+i)^(-n))/(i-g)}


fgivena<-function(a,i,n){pgivena(a,i,n)*(1+i)^n}
fgiveng<-function(g,i,n){pgiveng(g,i,n)*(1+i)^n}

PMT <- agivenp(10000,.1,3)
i1 <- 10000*.1
P1 <- PMT -i1
B1 <- 10000-P1

i2 <- B1*.1
P2 <- PMT -i2
B2 <- B1-P2

i3 <- B2*.1
P3 <- PMT - i3
B3 <- B2-P3


```

This is the amortization table for a \$10,000, 10\% per year loan with three annual payments.

Payment Number | Payment          | Interest  | Principal | Balance
---------------|---------         |---------- |-----------|--------
0              |                  |                 |                 |`r FinFormat(10000)`
1              |`r FinFormat(PMT)`|`r FinFormat(i1)`|`r FinFormat(P1)`|`r FinFormat(B1)`
2              |`r FinFormat(PMT)`|`r FinFormat(i2)`|`r FinFormat(P2)`|`r FinFormat(B2)`
3              |`r FinFormat(PMT)`|`r FinFormat(i3)`|`r FinFormat(P3)`|`r FinFormat(B3)`

+ Payments are given by $10000 (A|P, i = .1, 3) = `r Acc(agivenp(10000,.1,3))`$
+ Interest expense is the balance remaining times the effective interest rate per payment period.
+ Principal payment is the payment less the interest expense.
+ Balance remaining is the previous balance remaining less then principal payment.


## From the Loans Slides

```{r include=FALSE}
FinFormat <- function(x) format(x, digits=2, decimal.mark=".",   big.mark=",", small.interval=3)


Acc<-function(x){format(x,nsmall=2,digits=2,big.mark=',',scientific=FALSE)}
pgivena<-function(a,i,n){a*((1+i)^n -1)/(i*(1+i)^n)}
pgiveng<-function(g,i,n){g*( (1+i)^n -(i*n) -1  )/(i^2*(1+i)^n)}
agivenp<-function(p,i,n){p*(i*(1+i)^n)/((1+i)^n -1)}
pgivengrow<-function(a1,g,i,n){a1*(1-(1+g)^n*(1+i)^(-n))/(i-g)}


fgivena<-function(a,i,n){pgivena(a,i,n)*(1+i)^n}
fgiveng<-function(g,i,n){pgiveng(g,i,n)*(1+i)^n}

PMT <- agivenp(10000,.1,3)
i1 <- 10000*.1
P1 <- PMT -i1
B1 <- 10000-P1

i2 <- B1*.1
P2 <- PMT -i2
B2 <- B1-P2

i3 <- B2*.1
P3 <- PMT - i3
B3 <- B2-P3


```

Payment Number | Payment          | Interest  | Principal | Balance
---------------|---------         |---------- |-----------|--------
0              |                  |                 |                 |`r FinFormat(10000)`
1              |`r FinFormat(PMT)`|`r FinFormat(i1)`|`r FinFormat(P1)`|`r FinFormat(B1)`
2              |`r FinFormat(PMT)`|`r FinFormat(i2)`|`r FinFormat(P2)`|`r FinFormat(B2)`
3              |`r FinFormat(PMT)`|`r FinFormat(i3)`|`r FinFormat(P3)`|`r FinFormat(B3)`

+ Loan funding shows in the cash flow statement under finance as a positive number.
+ Interest expenses are shown on the income statement as a positive number.
+ Principal payments are show on the cash flow statement as a negative number.

##  In the Income and Cash Flow Statement


\begin{table}[h]
\begin{small}
\begin{tabular}{l||c|c|c|c|}
\multicolumn{5}{c}{Income Statement}\\
  &Year 0  &Year 1  & Year 2 & Year 3 \\
\hline
		Operating Revenue	& & 	& & \\
		Operating Expenses	&	&	& & \\
		Interest	&	&	1000&698 &366 \\
		Depreciation	&	&	& & \\
        \hline
		Taxable Income	&0	&(1000)	&(698) &(366) \\
		Tax	&	0 & (`r 1000*.4`)	&(`r 698*.4`) & (`r 366*.4`) \\
        \hline
		Net Income	& 0	&(`r NI1 <- 1000*.6;NI1`)	& (`r NI2 <- 698*.6; NI2`)& (`r NI3 <- 366*.6; NI3`)\\

\end{tabular}

\begin{tabular}{l||c|c|c|c|}
        \multicolumn{5}{c}{Cash Flow Statement}\\
  &Year 0  &Year 1  & Year 2 & Year 3 \\
\hline
	Operations&	&	& &\\
	\hspace{.25in}Net Income	& 0	& (`r NI1`) &(`r NI2`) &(`r NI3`)\\
	\hspace{.25in}Depreciation	&	&	& &\\
	Investments	&	&	& &\\
	Finance	&	&	& &\\
	\hspace{.25in}		Loan	&	10000 &	& &\\
	\hspace{.25in}		Loan	Payments&	&(3021)	&(3323) &(3656)\\
        \hline
		Net Cash Flow	& 10000	& (`r 600 + 3021`)	& (`r 418.8 + 3323`)& (`r 219.6 + 3656`)\\
\end{tabular}

\end{small}
\end{table}

## Working Capital Changes

These are common:

+ Increases in sales necessities a larger inventory (Working Capital)
+ More sales mean greater average invoices outstanding.

The logic is tricky only in the sense that you are showing the changes in these levels as cash flows.

## Example Inventory Changes

This shows what happens with average inventory changes from year-to-year.

Year            | 1   | 2   | 3   | 4   | 5   |
---             |---  |---  |---  |---  |---  |
Inventory       |1000 |1200 |900  |900  | 800 |
Working Capital | 0   |-200 |300  |0    |100  | 

+ Increases in inventory require more cash, outflows. 
+ Decreases free up cash, inflows.
+ Gets complicated when the current plan has inventory levels changing.  We will not do this (Difference in the differences).

## Summary

+ This is just a way of presenting the things that you have already done using accounting conventions.
+ Understandable by everyone that knows the conventions.
+ Only a few transactions:
    + Increase/Decrease in revenue
    + Increase/Decrease in sales
    + Asset purchase/sale with gains tax
    + Loan and loan repayment.
    + Working Capital Changes.
    
+ Worked Examples http://ec314-pdx-edu.wikidot.com/q4:income-and-cash-flow-statements