---
title: "Annual Worth"
output: beamer_presentation
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
options(scipen=999)

Acc<-function(x){format(x,nsmall=2,digits=2,big.mark=',',scientific=FALSE)}
pgivena<-function(a,i,n){a*((1+i)^n -1)/(i*(1+i)^n)}
pgiveng<-function(g,i,n){g*( (1+i)^n -(i*n) -1  )/(i^2*(1+i)^n)}
agivenp<-function(p,i,n){p*(i*(1+i)^n)/((1+i)^n -1)}
pgivengrow<-function(a1,g,i,n){a1*(1-(1+g)^n*(1+i)^(-n))/(i-g)}


fgivena<-function(a,i,n){pgivena(a,i,n)*(1+i)^n}
fgiveng<-function(g,i,n){pgiveng(g,i,n)*(1+i)^n}

```

## Goals

+ Present AW as a frequently preferred measure of costs and benefits.
+ Show details of AW calculation.
+ Give warnings about interpretation
+ Show a computational advantage in problems of repeated purchases.


## AW

$$AW(Asset) = PW(Asset)(A|P,i,Life~of~Asset)$$

+ AW is a transformation of present worth.
+ AW is, except in one edge case, smaller than PW.
    + The edge case is an asset that is installed at time zero and only lasts through time 1.
    + $(A|P,i,1) = \frac{i (1+i)}{ (1+i) -1} = (1+i)$
+ Many synonyms
    + Levelized cost
    + Capitalized cost (Danger PW is sometimes called this too)
    + Equivalent annual cost
    
## Simple Example

The car costs 10K and will last five years.  What is the per-period cost of the car when the MARR is 10%?

$$10K(A|P, i = 10\%, 5) = `r Acc(agivenp(10, .1, 5))`K$$

+ The purchase price -- now -- is a present worth.
+ Notice that it is not $\frac{10K}{5}$. It does take into account that you paid for the car in time zero.

## The AW Criteria

+ Annual Worth
    + Per-period costs and benefits ($)
    + Unconstrained: If $AW \geq 0$ get it.
    + Exclusive: If assets have the same life, choose asset with largest AW.
    
## Unconstrained Choice

If $AW \geq 0$ get it.

$$AW(Asset) = PW(Asset)(A|P,i,Life~of~Asset)$$

+ Note that $(A|P,i,Life~of~Asset)$ is always positive.
+ That means $PW \geq 0 \Rightarrow AW \geq 0$


## Exclusive Choice

If assets have the same life, choose asset with largest AW.

+ The "If" is critical
+ Example why

Year| A | B |
----|---|---|
0   | 0 | 0 |
1   |10 |9  |
2   |  |9  |
3   | |9  |
4   | |9  |
5   | |9  |
AW | 10 | 9|  

+ With moderate MARR, B is better but has lower AW.

## The If

+ It is less restrictive than you think and often useful.
+ You can construct assets from other assets.
+ Example {Car, Kayak Rack, Kayaks}
    + Doesn't make sense to get a rack without a car or kayaks.
    + Re-frame as exclusive choice: 
        + A = {Car}
        + B = {Car, Kayaks}
        + C = {Car, Rack, Kayaks}

## How this helps with repeated purchases

+ Remember the Roof Example?
    + Metal Roof (Like the old one but 'done correctly') would last 40 years and cost $1.5M.
    + PVC Roof would last 20 years and cost 800K to install.  It also requires 15K in additional inspection and maintenance in years 16-20.
+ These have unequal lives but you can use AW by constructing a new asset
    + A = {Metal now}
    + B = {PVC now, PVC in 20 years}
    
## The Metal Roof

What is the levelized cost of the $1.5M metal roof when the MARR is 10%?

## Answer

$$1500K(A|P, i = .1, 40) = `r Acc(agivenp(1500, .1, 40))`K$$

Interpret this as the annual cost of roofing services.

## The PVC Roof

What is the levelized cost of the the two PVC roofs when the MARR is 10%?

PW of the first roof is:
$$PW(PVC) = 800K + \frac{15K(P|A, i = .1, 5)}{(1+.1)^{15}} = `r Acc(800 + pgivena(15, .1, 5)/(1.1^15))`K$$

## Trick Question

The answer is the same if you make the calculation with two roofs or one.

```{r}
PW1 <- 800 + pgivena(15, .1, 5)/(1.1^15)
AW1 <- agivenp(PW1, .1, 20)
PW2 <- PW1 + PW1/1.1^20
AW2 <- agivenp(PW2, .1, 40)

```


+ One Roof: $ `r Acc(PW1)`K (A|P, i = 10\%, 20) = `r Acc(AW1)`K$
+ Two Roofs:
    + $PW(PVC| 40 Years) = `r Acc(PW1)`K + \frac{`r Acc(PW1)`K}{1.1^{20}} = `r Acc(PW2)`K$
    + $AW(PVC| 40 Years) = `r Acc(PW2)`K (A|P, i = 10\%, 40) = `r Acc(AW2)`K$
    
    

This is the computational advantage of AW in the case of repeated purchases.

## Repeated Purchases

+ AW provides a computational shortcut when there are repeated purchases.
    + If you buy one asset and then replace it with an identical one.
    + Like replacing lamps in fixtures.
+ Only works if the asset lives are factors of the planning horizon.
+ I usually call these conforming assets.
+ Example:  
    + Planning horizon of 12 years.
    + Options have lives of 3, 4, 2, and 6 years.
    
## Repeated Purchases (Con't)

As long as assets are factors of the planning horizon:

$$AW(Asset | Single) = AW (Asset| Repeated)$$

## The planning horizon

Often times you are not given an explicit planning horizon.

+ I've done this to you with abstract assets.  The planning horizon is implicitly the life of the longest lived asset.
+ Planning horizon is 2 years.
    
Year| A | B | 
----|---|---|
0   | 0 | 0 |
1   |10 |0  |
2   |0  |10 |

## Common Practice

+ Common practice is to set the planning horizon equal to the least common multiple of the asset lives.
    + Asset lives of 2, 3 and 4 years
    + Planning horizon of 12.
+ Same as result as infinite horizon without the limits.

## Comments on Common Practice

+ The planning horizon should be reasonable.
+ Backyard Aquaculture Story:
    + Two potential pumps: 3 and 4 year pumps
    + Planning horizon 12 years.
    + If 8 and 9 years?  I'm not living another 72 years.
+ Rounding a few years on asset lives often does not matter much for long-lived investments.
    + (P|A, i = 1., 100) = `r pgivena(1, .1, 100)`
    + (P|A, i = 1., 101) = `r pgivena(1, .1, 101)`
+ Do the math anyway, you are billing, but it probably won't change the decision.

## Common Practice  for Non-conforming Assets

+ Two general options:
    + Repeat purchases followed by an equipment lease for the remaining time.
    + Repeat purchases followed by salvage/sale of asset before end of life.
+ Calculate PW based on these patterns and convert to AW.
+ Example:  5 Year asset and 12 year planning horizon
    + Lease: $PW = PW(Asset) + \frac{PW(Asset)}{(1 +i)^{5} } + \frac{Lease(P|A, i, 2)}{(1+i)^{10}}$
    + Salvage: $PW = PW(Asset) + \frac{PW(Asset)}{(1 +i)^{5} } + \frac{PW(Asset)}{(1 +i)^{10} } +
    \frac{Salvage}{(1+i)^{12}}$
    
## Backyard Aquaculture

The idea was to add Tilapia to the composting and garden cycle. MARR of 10%

+ Choice of two pumps to keep water moving.
    + Pump 1: 7 year life, $100.
    + Pump 2: 5 year life, $75.
    
+ Calculate the AW
+ Choose the best asset to provide this service.

## Answer

+ Note implied planning horizon is 35 years.
+ AW Calculations
    + $AW(Pump 1) = 100 (A|P, i = .1, 7) = `r Acc(agivenp(100, .1, 7))`
    + $AW(Pump 2) = 75 (A|P, i = .1, 5) = `r Acc(agivenp(75, .1, 5))`
    
+ Interpretation
    + Pump 1 provides pumping services at a cost of \$`r Acc(agivenp(100, .1, 7))` per year.
    + Pump 2 provides pumping services at a cost of \$`r Acc(agivenp(75, .1, 5))` per year.


Notice that Pump 2 provides pumping services at the lowest cost.

## You Can Do PW too

PW Calculations for 35 Year planning horizon.

+ $PW(Pump~1| 35 Years) = 100 + \frac{100}{1.1^{7}}+ \frac{100}{1.1^{14}}+ \frac{100}{1.1^{21}} + \frac{100}{1.1^{28}} = `r Acc(100 + 100/1.1^7+ 100/1.1^14+ 100/1.1^21+ 100/1.1^28)`$
+ Will use the time compression trick.
```{r}
R5 <- round((1.1)^5-1,4)
```
    + Effective 5-year rate: $(1+\frac{.1}{1})^5 -1 = `r R5`$

    + $PW(Pump~2| 35 Years) = 75 + \frac{75(P|A, i = `r R5`, 6)}{(1+ `r R5`)} = `r Acc(75 + pgivena(75, R5, 6)/(1+ R5))`$
    
    
Notice that Pump 2 provides pumping services at the lowest cost.


## Summary

+ AW and PW give consistent choice but by different means.
+ AW provides a computational advantage when asset lives are factors of planning horizon.
+ Clients often find AW, per-period costs and benefits, easier to interpret than PW, life-cycle costs and benefits.


